Extended Data

### 2022-03-08 doc/manuscript/Dow_manuscript_Nature.Rmd
delete old content, misc improvements,  merge conflicted versions
#117

~~~~ (new intro content starting here)

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
However, model representation of C allocation to stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- and its seasonality is simplistic relative to *known decoupling of photosynthate production and stem growth [@delpierre_temperate_2016;@delpierre_wood_2016]*. <!-- go back and check that these references are appropriate for this statement-->
Given that the C allocated to woody growth has the longest residence time in the ecosystem[@refs], understanding whether warming spring temperatures will lead to sustained increases in forest C sequestration requires understanding the sensitivity of $ANPP_{woody}$ to spring temperatures.

We have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016]. 
Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@gao_earlier_2022;@anderson-teixeira_joint_2021; @more_conifer_refs]. 
Much less is known about deciduous tree responses[@savage_coordination_2021;@delpierre_temperate_2016].
Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].
However, earlier growth would not necessarily translate to earlier, faster, or more stem growth; rather, cell enlargement is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022]. 
Furthermore, *growth of broadleaf deciduous trees may be sink-saturated*, such that longer growing seasons with more C fixation do not necessarily augment growth[@refs].
Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms[@refs].

*points to make in new version*:

- stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
- stem growth and woody productivity are decoupled from GPP[@delpierre_wood_2016 and refs therein]

~~~~ ( old version starting here)

<!--These following two paragraphs have a lot of overlap and need work. They should probably be merged into one. -->
~~In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].~~
However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], *but rather can be limited by sink factors[@keenan_timing_2015;@zani_increased_2020;@zohner_how_2021;@delpierre_wood_2016]*,<!--I (KAT) need to work on the part in italics, if retained keenan_timing and zani_increased_2020 do not look at stem growth and are largely contradicted by lu_no_2022.--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]. 
~~These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.~~

~~While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented in the temperate deciduous biome and beyond [@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016] (but see Refs. [@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] for influence of spring temperatures on timing of growth in conifers).~~ 
The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021;@etzold_number_2022;@tumajer_limitation_2022;@zweifel_why_2021;@delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing and amount of growth remains unclear[@savage_coordination_2021]. <!--this sentence is repetitive with with previous.-->

*<PLACEHOLDER>* Three cambial processes drive the relationship between stem growth and woody productivity: cell differentiation, cell enlargement, and cell wall thickening. Irreversible stem expansion occurs as a result of differentiation (20% of stem growth) and expansion (80% of stem growth), and accounts for ~10% of annual woody productivity[@cuny_woody_2015]. The remaining 90% of woody productivity is caused by cell wall thickening. It would stand to reason that years with identical stem growth have a similar number of cells which are similarly sized (increased latewood can be an issue here, as you acknowledge). If we assume that cell wall thickness is controlled by processes within the tree (ie. hormonal responses instead of external climate drivers), then we can predict that cell wall thickness is relatively uniform within each cell type (early wood and late wood). Therefore, years with a similar amount of stem growth would sequester a similar amount of carbon during cell wall thickening, and we can say that stem growth equates to woody productivity.

~~Tree-ring records, which can be used to examine relationships of annual growth to temperature[@gao_earlier_2022] but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019]<!-- dorangeville_drought_2018 looks at PET.-->, with some hints that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019].<!-- add more refs here, but only if already cited elsewhere (to keep n references down-->
Thus, tree-ring evidence does not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake[@keenan_net_2014].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.~~

~~~~ (end old version)

>>>>

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons lead to sustained increases in annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
However, this assumption is overly simplistic given the known decoupling of photosynthate production and stem growth[@delpierre_temperate_2016;@delpierre_wood_2016], suggesting that model representation of C allocation to stem growth -- or woody aboveground net primary productivity ($ANPP_{woody}$) on the ecosystem level -- needs to be reworked[@delpierre_temperate_2016]. 
Given that the C allocated to woody growth has the longest residence time in the ecosystem[@refs], understanding whether warming spring temperatures will lead to sustained increases in forest C sequestration requires understanding the sensitivity of $ANPP_{woody}$ to spring temperatures.

We have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016]. 
Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@babst_twentieth_2019;@anderson-teixeira_joint_2021; @gao_earlier_2022]. 
Much less is known about deciduous tree responses[@savage_coordination_2021;@delpierre_temperate_2016].
Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].
However, earlier growth would not necessarily translate to earlier, faster, or more stem growth; rather, cell enlargement is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity, or _NEE_[@delpierre_wood_2016]. 
Furthermore, growth of broadleaf deciduous trees may be sink-saturated, such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

~~~*points to make*:

- stem growth is routinely used to estimate woody productivity (but wood density partially decouples them) [@delpierre_wood_2016 and refs therein]
- stem growth and woody productivity are decoupled from GPP[@delpierre_wood_2016 and refs therein]
- positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]
- The timing of stem growth and its climate sensitivity is linked to annual growth[@dorangeville_peak_2021;@etzold_number_2022;@tumajer_limitation_2022;@zweifel_why_2021;@delpierre_wood_2016]; however, the role of spring temperatures in shaping the timing and amount of growth remains unclear[@savage_coordination_2021].

~~~
